Ecosystem size and complexity dictate riverine biodiversity

### 2021-01-22 manuscript.Rmd
Update manuscript.Rmd

---
title: "Ecosystem size and complexity dictate riverine biodiversity"
author:
- Akira Terui^[Department of Biology, University of North Carolina at Greensboro]
- Seoghyun Kim\footnotemark[1]
- Christine L. Dolph^[Department of Ecology, Evolution and Behavior, University of Minnesota]
- Taku Kadoya^[National Institute for Environmental Studies]
- Yusuke Miyazaki^[Shiraume Gakuen University]
output:
  html_notebook: default
  word_document: default
abstract:
  The species-area relationship (the SAR), a tendency in which species richness increases with increasing area, is one of the most robust empirical generalizations in ecology and formed the basis of area-based conservation strategies. However, many ecosystems possess a complex spatial structure that cannot be represented by area (the scale-invariant complexity), and its role in scaling biodiversity is largely unknown. Here, we hypothesized that ecosystem size and complexity jointly regulate biodiversity in rivers. Rivers are a prime example of fractal branching networks where the recurrent merging of diverse streams generates a pronounced self-similarity in complex branching patterns. Ecosystem complexity, which we defined as the probability of branching per unit river distance, should increase watershed-scale species richness by increasing habitat heterogeneity (habitat diversity per unit area). Meanwhile, ecosystem size should dictate metacommunity size and total habitat diversity (area x habitat heterogeneity), two primary factors that regulate watershed-scale species richness. We developed theoretical models simulating metacommunity dynamics in branching river networks and yielded results consistent with our hypothesis. Ecosystem size and complexity increased watershed-scale species richness ($\gamma$ diversity), and the importance of spatial turnover in species composition ($\beta$ diversity) was apparent when simulated species dispersal is limited. Cross-watershed comparisons across two biogeographic regions (the Hokkaido island in Japan and the midwestern US) provided empirical support for the theoretical predictions. In both regions, watershed-scale species richness increased with ecosystem size (watershed area) and complexity (branching probability) through enhanced spatial turnover in species composition over the branching network. There is now a clear need to consider the dimension of ecosystem complexity in biodiversity conservation. The dual dimensional approach - size and complexity - should broaden viable options for spatial planning of protected areas or restoration sites, thereby helping reconcile the conflicts between biodiversity conservation and societal demands.

@@

Ecologists have long sought to understand the general drivers of biodiversity. One of the most robust empirical generalizations in ecology is the positive relationship between species richness and area, i.e., the species-area relationship (the SAR)[@lomolinoEcologyMostGeneral2000a]. In 1921, Arrhenius[@arrheniusSpeciesArea1921] formulated the SAR as a power-law $S = cA^z$ ($S$ is the number of species observed in a given geographic area $A$, $c$ a constant, and $z$ a scaling exponent), an equation currently known as the Arrhenius species-area relationship. Since then, the spatial scaling of species richness has been observed in many taxonomic groups[@triantisIslandSpeciesArea2012]. The SAR is ubiquitous because multiple mechanisms produce an apparently similar pattern. Larger ecosystems typically support more diverse metacommunities due to increased habitat diversity[@kallimanisHowDoesHabitat2008a], larger metacommunity size[@ben-hurDisentanglingMechanismsUnderlying2020;@hubbellUnifiedNeutralTheory2001], and/or enhanced colonization dynamics[@lososTheoryIslandBiogeography2009]. Importantly, the SAR provides the foundation for global conservation efforts. For example, conservation ecologists have used the SAR estimates to design marine and terrestrial protected areas, which currently encompass more than 30 million km^2^ globally[@deguignet2014UnitedNations2014].

Many ecosystems, however, possess a complex spatial structure that cannot be represented by area - a dimension referred to as scale-invariant complexity[@rinaldoEvolutionSelectionRiver2014; @rodriguez-iturbeFractalRiverBasins2001]. Such complexity is evident in branching ecosystems, including rivers, trees, and mountain ranges, to name just a few[@rodriguez-iturbeFractalRiverBasins2001]. Geomorphic or biological processes generate a pronounced self-similarity in complex branching patterns such that the part and the whole look alike[@rodriguez-iturbeFractalRiverBasins2001]. Even though the branching structure is independent of spatial scale, it forms a physical template that dictates habitat diversity and dispersal corridors for living organisms[@grantLivingBranchesPopulation2007; @rinaldoRiverNetworksEcological2018;@tonkinRoleDispersalRiver2018]. Limited, but accumulating evidence suggests that classical metapopulation and metacommunity theories cannot predict ecological dynamics driven by branching structure[@faganConnectivityFragmentationExtinction2002; @grantUseMultipleDispersal2010; @teruiMetapopulationStabilityBranching2018], and this recognition has led to recent developments of spatial theories devoted to complex branching ecosystems[@altermattDiversityRiverineMetacommunities2013; @carraraComplexInteractionDendritic2014; @carraraDendriticConnectivityControls2012; @teruiMetapopulationStabilityBranching2018]. However, most research has explored the consequences of branching complexity for local community structure[@altermattRiverNetworkProperties2013; @brownDendriticNetworkStructure2010] or has relied solely on theoretical arguments with limited replications of branching architecture[@holtRoleBranchingMaintenance2018]. At present, we lack a comprehensive evaluation of how branching complexity scales biodiversity at the metacommunity level. Filling this knowledge gap may provide common ground for achieving successful conservation in spatially complex ecosystems, where the accelerated species loss threatens the delivery of ecosystem services[@dudgeonFreshwaterBiodiversityImportance2006].

Here, we hypothesize that ecosystem size and complexity dictate biodiversity patterns in rivers, a prime example of complex branching ecosystems. The recurrent merging of diverse streams and rivers produces larger, ecologically distinct channels, ultimately forming a fluvial network with fractal branching patterns[@rodriguez-iturbeFractalRiverBasins2001]. As such, the complexity of branching structure, which we define here as the probability of branching per unit river distance [@teruiMetapopulationStabilityBranching2018; @yeakelSynchronisationStabilityRiver2014a], should control the habitat heterogeneity of the ecosystem (habitat diversity per unit area). Meanwhile, ecosystem size (watershed area) should determine the metacommunity size and total habitat diversity (area x heterogeneity). We predict that ecosystem size and branching complexity increase $\gamma$ diversity by enhancing either $\alpha$ and/or $\beta$ diversity under different ecological scenarios. The present study combines theory and empirical analysis of extensive community data from two different regions of the globe to provide crucial insights into how ecological communities are structured in complex branching networks.

First, we theoretically analyzed the influences of ecosystem size and branching complexity on $\gamma$ diversity. We depicted branching ecosystems as a spatial network of connected habitat patches[@teruiMetapopulationStabilityBranching2018; @yeakelSynchronisationStabilityRiver2014a] (**Figure 1**) where two factors determine mean environmental conditions at each habitat patch: (1) headwater environments (the most upstream habitat patch) and (2) local environmental noise. Environmental values at headwaters are drawn randomly from a normal distribution and propagate downstream with local environmental noise (i.e., a spatial autoregressive process with white noise). These environmental values recurrently 'mix' at confluences considering the relative size of joining tributaries (see Methods). Therefore, our network-generation procedure resembles natural processes of how branching river networks create diverse habitats in a metacommunity. We then simulated metacommunity dynamics in the theoretical branching networks using a general metacommunity framework[@thompsonProcessbasedMetacommunityFramework2020]. In this model, 50 species with different abiotic niches (optimum and width) disperse along a network and compete for resources with varied strengths in temporally dynamic environments (see Methods for model details).

We considered `r n_scenario` simulation scenarios, which comprised a combination of four landscape and eight ecological scenarios. We distinguished landscape scenarios by setting four combinations of environmental variation at headwaters ($\sigma_h =$ `r unique(dat_sim$sd_env_source)`) and the degree of local environmental noise ($\sigma_l =$ `r unique(dat_sim$sd_env_lon)`). When $\sigma_h > \sigma_l$, branching produces greater habitat heterogeneity because headwaters are the primary source of environmental variation. In the meantime, when $\sigma_h \le \sigma_l$, local environmental noise masks environmental variation among tributaries, leading to minimal influences of branching on habitat heterogeneity. Thus, the inequality between $\sigma_h$ and $\sigma_l$ creates contrasting patterns of habitat heterogeneity withing a network. We also considered eight ecological scenarios that differ in dispersal distance (controlled by the rate parameter of an exponential dispersal kernel), dispersal probability, and the maximum value of interspecific competition strength (see Methods). Our simulation scenarios were capable of reproducing common spatial patterns of local species richness, corroborating the appropriateness of our choice in parameter combinations (Supplementary Information). Under each simulation scenario, we simulated 1000 time steps of metacommunity dynamics in `r max(dat_sim$n_rep)` branching networks with the gradients of ecosystem size (the number of habitat patches: `r min(dat_sim$n_patch)` to `r max(dat_sim$n_patch)`) and complexity (branching probability: `r round(min(dat_sim$p_branch), 2)` to `r round(max(dat_sim$p_branch), 2)`).

@@

We depicted branching ecosystems as a spatial network of connected habitat patches[@teruiMetapopulationStabilityBranching2018; @yeakelSynchronisationStabilityRiver2014a]. Habitat patches can be either non-branching or branching river sections with a unit length $l$, which defines the scale of local species interactions. Two parameters determine the geometric properties of simulated branching networks: the number of habitat patches $N_p$ (ecosystem size) and branching probability $P_b$ (ecosystem complexity). Each habitat patch is assigned to be a branching patch (including upstream terminals) with probability $P_b$ or a non-branching patch with probability $1-P_b$. In this framework, an individual branch is a consecutive series of non-branching patches terminated at a branching patch; therefore, the number of habitat patches in a single branch $n_b$ is a realization of a random variable drawn from a geometric distribution $n_{b} \sim Ge(P_{b})$. This representation has two merits. First, it reflects observed patterns of branch length distribution, which is known to follow an exponential distribution (a continuous version of a geometric distribution). Second, it preserves the fractal nature of branching patterns[@peckhamReformulationHortonLaws1999].

Two sources of variation characterize the long-term average of abiotic environment $\mu_z$ at each habitat patch: environmental variation at headwaters ($\sigma_h$) and local environmental noise ($\sigma_l$). We draw random values from a normal distribution with a mean of zero and SD of $\sigma_{h}$ and assigned them to headwater patches (i.e., the most upstream patches). These environmental values at headwaters propagate downstream through a spatial autoregressive process defined as:

@@

where $\mu_{z,k}$ is the environmental value at longitudinal position $k$ ($k$ is the network distance from the outlet patch; $k = 1$ at the outlet), $\rho$ is the strength of spatial autocorrelation and $\epsilon_k$ is the local environmental noise that follows a normal distribution with a mean of zero and SD of $\sigma_l$. The parameter $\rho$ can take values of $0-1$ with larger values indicating greater spatial autocorrelation. In this study, we set $\rho = 1$ to mimic strong spatial autocorrelation in riverine environments. At confluences, we took a weighted mean of environmental values given the relative size of upstream contributing area of joining tributaries:

$$
\begin{align}
\mu_{z,k} &= \omega(\rho \mu_{z,k+1}^R + \epsilon_{k}^R) + (1-\omega)(\rho \mu_{z,k+1}^L + \epsilon_{k}^L) \\
\omega &= \frac{A_{k+1}^R}{A_{k+1}^R + A_{k+1}^L}
\end{align}
$$
where $R$ and $L$ denote "right" and "left" branches, respectively, and $A_k$ is the number of upstream habitat patches at longitudinal position $k$ (akin to the upstream watershed area). With this expression, larger tributaries have a greater influence on the downstream environment, as observed in natural river networks[@bendaNetworkDynamicsHypothesis2004].

@@

The environmental value at patch $x$ and time $t$, $z_x(t)$, is assumed to follow a multivariate normal distribution as $z_{x}(t) \sim MVN(\mu_z, \Omega_z)$. $\mu_z$ is the vector of mean environmental conditions of patches, and $\Omega_z$ is the variance-covariance matrix. Spatial autocorrelation in temporal environmental dynamics is considered by expressing the off-diagonal elements as:

@@

We allowed interspecific variation in niche optimum and width, which were drawn from uniform distributions as $\mu_i \sim Unif(-1, 1)$ and $\sigma_{niche,i} \sim Unif(0.1, 1)$. We used fixed values for the following parameters: maximum reproductive number ($r_{0,i} =$ 4), niche cost ($\nu =$ `r unique(dat_sim$niche_cost)`), the degree of temporal fluctuation in abiotic environments ($\sigma_z =$ `r unique(dat_sim$sd_env)`), and the degree of spatial autocorrelation in temporal environmental dynamics ($\phi =$ `r unique(dat_sim$phi)`).

Under each simulation scenario, we simulated 1400 time steps of metacommunity dynamics (including 200 time steps for initialization and 200 time steps for burn-in) in `r max(dat_sim$n_rep)` branching networks with the gradients of ecosystem size (the number of habitat patches: `r min(dat_sim$n_patch)` to `r max(dat_sim$n_patch)`) and complexity (branching probability: `r round(min(dat_sim$p_branch), 2)` to `r round(max(dat_sim$p_branch), 2)`). During the first 200 time steps (an initialization period), we initialized the simulation by seeding each habitat patch with populations of each species drawn from a Poisson distribution with a mean of 0.5. We repeated the seeding every 10 time steps during the initialization to allow species to establish their populations if they can grow from low abundances. After the initialization period, we ran 200 time steps as a burn-in period with no seeding to minimize influences of initial conditions. We saved the last 1000 time steps to estimate the temporal averages of $\alpha$, $\beta$, and $\gamma$ diversity. We removed simulation replicates in which no species establish populations over the initial 400 time steps (`r sum(dat_sim$alpha_div == 0)` replicates out of `r nrow(dat_sim)` replicates). Ecosystem size and complexity were drawn randomly from uniform distributions as: $N_p \sim Unif(10, 150)$ and $P_b \sim Unif(0.01, 0.99)$ ($N_p$ was rounded to the nearest integer before running simulations).

@@

We compiled fish data at two biogeographic regions: the Hokkaido island, Japan and the Midwest, US. For Hokkaido, we used data from the Hokkaido Freshwater Fish Database HFish, monitoring data at protected watersheds, and primary data collected from literature, which collectively cover the entire Hokkaido island. For the Midwest, we assembled fish community data collected by Iowa Department of Natural Resources, Illinois Environmental Protection Agency and Illinois Department of Natural Resources, Minnesota Pollution Control Agency, and Wisconsin Department of Natural Resources. We detailed procedures for initial data selection in Supplementary Information.

@@

We estimated $\alpha$, $\beta$, and $\gamma$ diversity at each watershed. $\gamma$ diversity is the total species richness at each watershed. Since it is impossible to sample all species present in each watershed, we used asymptotic species richness as a proxy for $\gamma$ diversity. Asymptotic species richness can be interpreted as the estimated true species richness in a given area when estimated using spatial replicates of local community data within a metacommunity[@chaoRarefactionExtrapolationHill2014a; @chaseEcosystemDecayExacerbates2020]. We estimated asymptotic species richness using sampling sites located in the same watershed (i.e., spatial replicates within a metacommunity) via the R package iNEXT[@hsiehINEXTPackageRarefaction2016]. We assessed the completeness of the observed species richness based on sample coverage, which is the proportion of observed species richness to asymptotic species richness. We excluded watersheds whose sample coverage was below 0.9 because the estimates of asymptotic species richness may contain non-negligible statistical uncertainty. $\alpha$ diversity was estimated by taking an average of local species richness at each site. $\beta$ diversity was estimated by dividing the estimated $\gamma$ diversity by $\alpha$ diversity.

@@

We collected data for watershed characteristics (watershed area and branching probability), climatic variables (annual mean air temperature and cumulative precipitation), and land use (% forest). We delineated watershed polygons and river lines (defined as $\ge$ 1 km^2^) using Hydrologically Adjusted Elevations of MERIT Hydro[@yamazakiMERITHydroHighResolution2019] with ArcMap 10.7 and SAGA 6.3.0 combined with the R package 'RSAGA'[@brenningRSAGASAGAGeoprocessing2018]. Watershed area was estimated based on the delineated watershed polygons using functions in the R package 'sf'[@pebesmaSimpleFeaturesStandardized2018]. Branching probability was estimated by fitting exponential distributions to histograms of branch length (unit: kilometer, river segment between successive confluences or a confluence and the outlet/upstream terminal) as $branch~length \sim Exp(\lambda)$. A derived statistical quantity $1 − e^{−\lambda}$ corresponds to the theoretical branching probability $P_b$. We estimated the spatial averages of air temperature and cumulative precipitation at a 1-km resolution for each watershed, which were extracted from WorldClim 1.4[@hijmansVeryHighResolution2005] via the R package 'raster.' We calculated % forest for each watershed using land-use data in 2015 from Copernicus Global Land Service (100-m resolution)[@marcelbuchhornCopernicusGlobalLand2020].

@@

We developed linear models to examine the effects of macro-scale drivers on diversity metrics. Log-transformed diversity metrics ($\alpha$, $\beta$ and $\gamma$ diversity) at watershed $i$, $log_{10}~y_i$, are assumed to follow a normal distribution as $log_{10}~y_i \sim Normal(\mu_i, \sigma^2)$. The parameter $\mu_i$ was related to linear predictors, including log-transformed watershed area (ecosystem size; $log_{10}~WA$), log-branching probability (ecosystem complexity; $log_{10}~P_b$), region (coded as Hokkaido = 0, Midwest = 1; $region$), air temperature, precipitation, and percent of forest land use (logit-transformed). Although our primary focus of this analysis is to assess the effects of watershed area and branching probability, we included climate and land-use variables to control these effects statistically.

>>>>

---
title: "Ecosystem size and complexity dictate riverine biodiversity"
author:
- Akira Terui^[Department of Biology, University of North Carolina at Greensboro]
- Seoghyun Kim\footnotemark[1]
- Christine L. Dolph^[Department of Ecology, Evolution and Behavior, University of Minnesota]
- Taku Kadoya^[National Institute for Environmental Studies]
- Yusuke Miyazaki^[Shiraume Gakuen University]
output:
  html_notebook: default
  word_document: default
abstract:
  The species-area relationship (the SAR), a tendency in which species richness increases with increasing area, is one of the most robust empirical generalizations in ecology and formed the basis of area-based conservation strategies. However, many ecosystems possess a complex spatial structure that cannot be represented by area (the scale-invariant complexity), and its role in scaling biodiversity is largely unknown. Here, we hypothesized that ecosystem size and complexity jointly regulate biodiversity in rivers. Rivers are a prime example of fractal branching networks where the recurrent merging of diverse streams generates a pronounced self-similarity in complex branching patterns. Ecosystem complexity, which we defined as the probability of branching per unit river distance, should increase watershed-scale species richness by increasing habitat heterogeneity (habitat diversity per unit area). Meanwhile, ecosystem size should dictate metacommunity size and total habitat diversity (area x habitat heterogeneity) that regulate watershed-scale species richness. We developed theoretical models simulating metacommunity dynamics in branching river networks and yielded results consistent with our hypothesis. Ecosystem size and complexity increased watershed-scale species richness ($\gamma$ diversity), and the importance of spatial turnover in species composition ($\beta$ diversity) was apparent when simulated species dispersal is limited. Cross-watershed comparisons across two biogeographic regions (the Hokkaido island in Japan and the midwestern US) provided empirical support for the theoretical predictions. In both regions, watershed-scale species richness increased with ecosystem size (watershed area) and complexity (branching probability) through enhanced spatial turnover in species composition over the branching network. There is now a clear need to consider the dimension of ecosystem complexity in biodiversity conservation. The dual dimensional approach - size and complexity - should broaden viable options for spatial planning of protected areas or restoration sites, thereby helping reconcile the conflicts between biodiversity conservation and societal demands.

@@

Ecologists have long sought to understand the general drivers of biodiversity. One of the most robust empirical generalizations in ecology is the positive relationship between species richness and area, i.e., the species-area relationship (the SAR)[@lomolinoEcologyMostGeneral2000a]. In 1921, Arrhenius[@arrheniusSpeciesArea1921] formulated the SAR as a power-law $S = cA^z$, an equation currently known as the Arrhenius species-area relationship ($S$ is the number of species observed in a given geographic area $A$, $c$ the constant, and $z$ the scaling exponent). Since then, the spatial scaling of species richness has been observed in many taxonomic groups[@triantisIslandSpeciesArea2012]. The SAR is ubiquitous because multiple mechanisms produce an apparently similar pattern. Larger ecosystems typically support more diverse metacommunities due to increased habitat diversity[@kallimanisHowDoesHabitat2008a], larger metacommunity size[@ben-hurDisentanglingMechanismsUnderlying2020;@hubbellUnifiedNeutralTheory2001], and/or enhanced colonization dynamics[@lososTheoryIslandBiogeography2009]. Importantly, the SAR provides the foundation for global conservation efforts. For example, conservation ecologists have used the SAR estimates to design marine and terrestrial protected areas, which currently encompass more than 30 million km^2^ globally[@deguignet2014UnitedNations2014].

Many ecosystems, however, possess a complex spatial structure that cannot be represented by area - a dimension referred to as scale-invariant complexity[@rinaldoEvolutionSelectionRiver2014; @rodriguez-iturbeFractalRiverBasins2001]. Such complexity is evident in branching ecosystems, including rivers, trees, and mountain ranges, to name just a few[@rodriguez-iturbeFractalRiverBasins2001]. Geomorphic or biological processes generate a pronounced self-similarity in complex branching patterns such that the part and the whole look alike[@rodriguez-iturbeFractalRiverBasins2001]. Even though the branching structure is independent of spatial scale, it forms a physical template that dictates habitat diversity and dispersal corridors for living organisms[@grantLivingBranchesPopulation2007; @rinaldoRiverNetworksEcological2018;@tonkinRoleDispersalRiver2018]. Limited, but accumulating evidence suggests that classical metapopulation and metacommunity theories cannot predict ecological dynamics driven by branching structure[@faganConnectivityFragmentationExtinction2002; @grantUseMultipleDispersal2010; @teruiMetapopulationStabilityBranching2018], and this recognition has led to recent developments of spatial theories devoted to complex branching ecosystems[@altermattDiversityRiverineMetacommunities2013; @carraraComplexInteractionDendritic2014; @carraraDendriticConnectivityControls2012; @teruiMetapopulationStabilityBranching2018]. However, most research has explored the consequences of branching complexity for local community structure[@altermattRiverNetworkProperties2013; @brownDendriticNetworkStructure2010] or has relied solely on theoretical arguments with limited replications of branching architecture[@holtRoleBranchingMaintenance2018]. At present, we lack a comprehensive evaluation of how branching complexity scales biodiversity at the metacommunity level. Filling this knowledge gap may provide common ground for achieving successful conservation in spatially complex ecosystems, where the accelerated species loss threatens the delivery of ecosystem services[@dudgeonFreshwaterBiodiversityImportance2006a].

Here, we hypothesize that ecosystem size and complexity dictate biodiversity patterns in rivers, a prime example of complex branching ecosystems. The recurrent merging of diverse streams and rivers produces larger, ecologically distinct channels, ultimately forming a fluvial network with fractal branching patterns[@rodriguez-iturbeFractalRiverBasins2001]. As such, the complexity of branching structure, which we define here as the probability of branching per unit river distance [@teruiMetapopulationStabilityBranching2018; @yeakelSynchronisationStabilityRiver2014], should control the ecosystem's habitat heterogeneity (habitat diversity per unit area). Meanwhile, ecosystem size (watershed area) should determine the metacommunity size and total habitat diversity (area x heterogeneity). We predict that ecosystem size and branching complexity increase watershed-scale species richness ($\gamma$ diversity) by enhancing local species richness ($\alpha$ diversity) and/or spatial turnover of species composition ($\beta$ diversity) under different ecological scenarios. The present study combines theory and empirical analysis of extensive community data from two different regions of the globe to provide crucial insights into how ecological communities are structured in complex branching networks.

First, we theoretically analyzed the influences of ecosystem size and branching complexity on $\gamma$ diversity. We depicted branching ecosystems as a spatial network of connected habitat patches or local communities[@teruiMetapopulationStabilityBranching2018; @yeakelSynchronisationStabilityRiver2014] (**Figure 1**) where two factors determine mean environmental conditions at each habitat patch: (1) headwater environments (the most upstream habitat patch) and (2) local environmental noise. Environmental values at headwaters are drawn randomly from a normal distribution and propagate downstream with local environmental noise (i.e., a spatial autoregressive process with white noise). These environmental values recurrently 'mix' at confluences considering the relative size of joining tributaries (see Methods). Therefore, our network-generation procedure resembles natural processes of how branching river networks create diverse habitats in a metacommunity. We then simulated metacommunity dynamics in the theoretical branching networks using a general metacommunity framework[@thompsonProcessbasedMetacommunityFramework2020]. In this model, 50 species with different abiotic niches (optimum and width) disperse along a network and compete for resources with varied strengths in temporally dynamic environments (see Methods for model details).

We considered `r n_scenario` simulation scenarios, which comprised a combination of four landscape and eight ecological scenarios. We distinguished landscape scenarios by setting four combinations of environmental variation at headwaters ($\sigma_h =$ `r unique(dat_sim$sd_env_source)`) and the degree of local environmental noise ($\sigma_l =$ `r unique(dat_sim$sd_env_lon)`), both of which are defined as standard deviations of normal distributions. When $\sigma_h > \sigma_l$, branching produces greater habitat heterogeneity because headwaters are the primary source of environmental variation. In the meantime, when $\sigma_h \le \sigma_l$, local environmental noise masks environmental variation among tributaries, leading to minimal influences of branching on habitat heterogeneity. Thus, the inequality between $\sigma_h$ and $\sigma_l$ creates contrasting patterns of habitat heterogeneity withing a network. We also considered eight ecological scenarios that differ in dispersal distance (controlled by the rate parameter of an exponential dispersal kernel), dispersal probability, and the maximum value of interspecific competition strength (see Methods). These simulation scenarios were capable of reproducing common spatial patterns of local species richness in rivers, corroborating the appropriateness of our choice in parameter combinations (Supplementary Information). Under each simulation scenario, we simulated 1000 time steps of metacommunity dynamics in `r max(dat_sim$n_rep)` branching networks with the gradients of ecosystem size (the number of habitat patches: `r min(dat_sim$n_patch)` to `r max(dat_sim$n_patch)`) and complexity (branching probability: `r round(min(dat_sim$p_branch), 2)` to `r round(max(dat_sim$p_branch), 2)`).

@@

We depicted branching ecosystems as a bifurcating spatial network of connected habitat patches[@teruiMetapopulationStabilityBranching2018; @yeakelSynchronisationStabilityRiver2014] (**Figure 1**). Habitat patches, or local communities, can be either non-branching or branching river sections with a unit length $l$, which defines the scale of local species interactions. Two parameters determine the geometric properties of simulated branching networks: the number of habitat patches $N_p$ (ecosystem size) and branching probability $P_b$ (ecosystem complexity). Each habitat patch is assigned to be a branching patch (including upstream terminals) with probability $P_b$ or a non-branching patch with probability $1-P_b$. In this framework, an individual branch is a consecutive series of non-branching patches terminated at a branching patch; therefore, the number of habitat patches in branch $q$, $n_{p,q}$, is a realization of a random variable drawn from a geometric distribution $n_{p,q} \sim Ge(P_{b})$[@teruiMetapopulationStabilityBranching2018] but conditional on $\sum_{q=1}^{N_b}n_{p,q} = N_p$ ($N_b$ is the number of branches in a network). The number of branches $N_b$ is identical to the number of branching habitat patches in a network, or the number of 'success' out of $N_p$ trials. This leads to an assumption that $N_b$ is a realization of a random variable drawn from a binomial distribution as $N_b \sim Binomial(N_p, P_b)$. This framework has two merits. First, it reflects observed patterns of branch length distribution, which is known to follow an exponential distribution (a continuous version of a geometric distribution)[@rodriguez-iturbeFractalRiverBasins2001]. Second, it preserves the fractal nature of branching patterns[@peckhamReformulationHortonLaws1999].

In our network generation, we first drew $N_b$ from the binomial distribution but selected an odd number to constitute a bifurcating branching network. We then determined $n_{p,q}$ as random draws from the geometric distribution with the constraint of $\sum_{q=1}^{N_b}n_{p,q} = N_p$. The generated branches were organized randomly into a bifurcating branching network.

Two sources of variation characterize the long-term average of abiotic environment at each habitat patch: environmental variation at headwaters ($\sigma_h$) and local environmental noise ($\sigma_l$). We draw random values from a normal distribution with a mean of zero and SD of $\sigma_{h}$ and assigned them to headwater patches (i.e., the most upstream patches). These environmental values at headwaters propagate downstream through a spatial autoregressive process defined as:

@@

where $\mu_{z,k}$ is the mean environmental value at longitudinal position $k$ ($k$ is the network distance from the outlet patch; $k = 1$ at the outlet), $\rho$ is the strength of spatial autocorrelation and $\epsilon_k$ is the local environmental noise that follows a normal distribution with a mean of zero and SD of $\sigma_l$. The parameter $\rho$ can take values of $0-1$ with larger values indicating greater spatial autocorrelation. In this study, we set $\rho = 1$ to mimic strong spatial autocorrelation in riverine environments. At confluences, we took a weighted mean of environmental values given the relative size of upstream contributing area of joining tributaries:

$$
\begin{align}
\mu_{z,k} &= \omega(\rho \mu_{z,k+1}^R + \epsilon_{k}^R) + (1-\omega)(\rho \mu_{z,k+1}^L + \epsilon_{k}^L) \\
\omega &= \frac{A_{k+1}^R}{A_{k+1}^R + A_{k+1}^L}
\end{align}
$$
where $R$ and $L$ denote "right" and "left" branches, respectively, and $A_k$ is the number of upstream habitat patches at longitudinal position $k$ (akin to the upstream watershed area). The parameter $\omega$ takes values of $0-1$ and represents the relative influence of the right tributary given the relative size of joining tributaries. With this expression, larger tributaries have a greater influence on the downstream environment, as observed in natural river networks[@bendaNetworkDynamicsHypothesis2004].

@@

The environmental value at patch $x$ and time $t$, $z_x(t)$, is assumed to follow a multivariate normal distribution as $z_{x}(t) \sim MVN(\mu_z, \Omega_z)$. $\mu_z$ is the vector of mean environmental conditions of habitat patches, and $\Omega_z$ is the variance-covariance matrix. Spatial autocorrelation in temporal environmental dynamics is considered by expressing the off-diagonal elements as:

@@

We allowed interspecific variation in niche optimum and width, which were drawn from uniform distributions as $\mu_i \sim Unif(-1, 1)$ and $\sigma_{niche,i} \sim Unif(0.1, 1)$. We used fixed values for the following parameters: maximum reproductive rate ($r_{0,i} =$ 4), niche cost ($\nu =$ `r unique(dat_sim$niche_cost)`), the degree of temporal fluctuation in abiotic environments ($\sigma_z =$ `r unique(dat_sim$sd_env)`), and the degree of spatial autocorrelation in temporal environmental dynamics ($\phi =$ `r unique(dat_sim$phi)`).

Under each simulation scenario, we simulated 1400 time steps of metacommunity dynamics (including 200 time steps for initialization and 200 time steps for burn-in) in `r max(dat_sim$n_rep)` branching networks with the gradients of ecosystem size (the number of habitat patches: `r min(dat_sim$n_patch)` to `r max(dat_sim$n_patch)`) and complexity (branching probability: `r round(min(dat_sim$p_branch), 2)` to `r round(max(dat_sim$p_branch), 2)`). During the first 200 time steps (an initialization period), we initialized the simulation by seeding each habitat patch with populations of each species drawn from a Poisson distribution with a mean of 0.5. We repeated the seeding every 10 time steps during the initialization to allow species to establish their populations if they can grow from low abundances. After the initialization period, we ran 200 time steps as a burn-in period with no seeding to minimize influences of initial conditions. We saved the last 1000 time steps to estimate the temporal averages of $\alpha$, $\beta$, and $\gamma$ diversity. We removed simulation replicates in which no species established populations over the initial 400 time steps (`r sum(dat_sim$alpha_div == 0)` replicates out of `r nrow(dat_sim)` replicates). Ecosystem size and complexity were drawn randomly from uniform distributions as: $N_p \sim Unif(10, 150)$ and $P_b \sim Unif(0.01, 0.99)$ ($N_p$ was rounded to the nearest integer before running simulations).

@@

We compiled fish data at two biogeographic regions: the Hokkaido island, Japan and the Midwest, US. For Hokkaido, we used data from the Hokkaido Freshwater Fish Database HFish[@fukushimaHokkaidoFreshwaterFish2011; @fukushimaModellingEffectsDams2007], monitoring data at protected watersheds[@comteRivFishTIMEGlobalDatabase2021; @teruiMetapopulationStabilityBranching2018], and primary data collected by the authors[@miyazakiTemporalDynamicsFluvial2016; @teruiThreeEcologicalFactors2016], which collectively cover the entire Hokkaido island. For the Midwest, we assembled fish community data collected by Iowa Department of Natural Resources, Illinois Environmental Protection Agency and Illinois Department of Natural Resources, Minnesota Pollution Control Agency, and Wisconsin Department of Natural Resources. We detailed procedures for initial data selection in Supplementary Information.

@@

We estimated $\alpha$, $\beta$, and $\gamma$ diversity at each watershed. $\gamma$ diversity is the total species richness at each watershed. Since it is impossible to sample all species present in each watershed, we used asymptotic species richness as a proxy for $\gamma$ diversity. Asymptotic species richness can be interpreted as the estimated true species richness in a given area when estimated using spatial replicates of local community data within a metacommunity[@chaoRarefactionExtrapolationHill2014; @chaseEcosystemDecayExacerbates2020]. We estimated asymptotic species richness using sampling sites located in the same watershed (i.e., spatial replicates within a metacommunity) via the R package iNEXT[@hsiehINEXTPackageRarefaction2016]. We assessed the completeness of the observed species richness based on sample coverage, which is the proportion of observed species richness to asymptotic species richness. We excluded watersheds whose sample coverage was below 0.9 because the estimates of asymptotic species richness may contain non-negligible statistical uncertainty. $\alpha$ diversity was estimated by taking an average of local species richness at each site. $\beta$ diversity was estimated by dividing the estimated $\gamma$ diversity by $\alpha$ diversity.

@@

We collected data for watershed characteristics (watershed area and branching probability), climatic variables (annual mean air temperature and cumulative precipitation), and land use (% forest). We delineated watershed polygons and river lines (defined as $\ge$ 1 km^2^) using Hydrologically Adjusted Elevations of MERIT Hydro[@yamazakiMERITHydroHighResolution2019] with ArcMap 10.7 and SAGA 6.3.0 combined with the R package 'RSAGA'[@brenningRSAGASAGAGeoprocessing2018]. Watershed area was estimated based on the delineated watershed polygons using functions in the R package 'sf'[@pebesmaSimpleFeaturesStandardized2018]. Branching probability was estimated by fitting exponential distributions to histograms of branch length (unit: kilometer, river segment between successive confluences or a confluence and the outlet/upstream terminal) as $branch~length \sim Exp(\lambda)$. A derived statistical quantity $1 − e^{−\lambda}$ corresponds to the theoretical branching probability $P_b$. We estimated the spatial averages of air temperature and cumulative precipitation at a 1-km resolution for each watershed, which were extracted from WorldClim 1.4[@hijmansVeryHighResolution2005] via the R package 'raster.' We calculated % forest for each watershed using land use data in 2015 from Copernicus Global Land Service (100-m resolution)[@marcelbuchhornCopernicusGlobalLand2020].

@@

We developed linear models to examine the effects of macro-scale drivers on diversity metrics. Log-transformed diversity metrics ($\alpha$, $\beta$ and $\gamma$ diversity) at watershed $i$, $log_{10}~y_i$, are assumed to follow a normal distribution as $log_{10}~y_i \sim Normal(\mu_i, \sigma^2)$. The parameter $\mu_i$ was related to linear predictors, including log-transformed watershed area (ecosystem size; $log_{10}~WA$), log-branching probability (ecosystem complexity; $log_{10}~P_b$), region (coded as Hokkaido = 0, Midwest = 1; $region$), air temperature, precipitation, and percent of forest land use (logit-transformed). Although our primary focus of this analysis is to assess the effects of watershed area and branching probability, we included climate and land use variables to control these effects statistically.
